---
title: "Exploratory Analysis"
output: html_notebook
---

```{r, warning=FALSE,message=FALSE}
pacman::p_load(
  tidyverse,
  here,
  RColorBrewer,
  lubridate,
  scales,
  GGally,
  stats,
  corrplot,
  leaps,
  glmulti,
  broom,
  rpart,
  rpart.plot,
  modelr,
  yardstick,
  caret,
  ranger
)
```

```{r, warning=FALSE,message=FALSE}
flights_dt <- read_csv(here("clean_data/flights_clean.csv"))

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

summary(flights_dt)
```

# Which airport in our dataset has the highest number of departures?

```{r}
flights_dt %>% 
  group_by(origin) %>%
  ggplot() +
  aes(x = origin, fill = origin) +
  geom_bar() +
  scale_fill_manual(values = cbPalette) +
  labs(
    x = "airport",
    y = "departure numbers",
    title = "Departure numbers by airport"
  ) +
  guides(fill = "none") +
  theme_bw()
```

# What is the trend of delays over the year, how does it compare across the three airports?

```{r,warning=FALSE}
flights_dt %>% 
  filter(dep_delay > 0) %>% 
  group_by(month = floor_date(sched_dep_time, "month"), origin) %>% 
  summarise(mean_delay = mean(dep_delay), .groups = "drop") %>% 
  ggplot() +
  aes(x = month, y = mean_delay, colour = origin) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.8) +
  labs(
    x = "month",
    y = "mean delay (minutes)",
    title = " Mean delay by month",
    colour = "airport"
  ) +
  scale_colour_manual(values = c("#999999", "#E69F00", "#56B4E9")) +
  theme_bw()
```

# What is the number of departure delays over the year?

```{r}
flights_dt %>%
  filter(dep_delay > 0) %>% 
  group_by(month = floor_date(sched_dep_time, "month"), origin) %>% 
  summarise(no_of_delays = n(), .groups = "drop") %>% 
  ggplot() +
  aes(x = month, y = no_of_delays, colour = origin) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.8) +
  labs(
    x = "month",
    y = "number of delays",
    title = " Monthly trend of departure delays",
    colour = "airport"
  ) +
  scale_colour_manual(values = c("#999999", "#E69F00", "#56B4E9")) +
  theme_bw()
```

# What is the number of flights per month?

```{r}
flights_dt %>%
  mutate(month = month(sched_dep_time, label = TRUE)) %>% 
  group_by(month, origin) %>% 
  summarise(no_of_flights = n(), .groups = "drop") %>% 
  ggplot() +
  aes(x = month, y = no_of_flights, fill = origin) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(
    x = "month",
    y = "number of departures",
    title = "Departure numbers by month",
    fill = "airport"
  ) +
  scale_fill_manual(values = cbPalette) +
  theme_bw()
```

# Which carrier has the most departure delays?

```{r}
flights_dt %>% 
  filter(dep_delay > 0) %>% 
  group_by(carrier_name, origin) %>% 
  summarise(mean_delay = mean(dep_delay), .groups = "drop") %>% 
  ggplot() +
  aes(x = reorder(carrier_name, mean_delay), y = mean_delay, fill = origin) +
  geom_col(alpha = 0.8) +
  facet_wrap(~ origin) +
  labs(
    x = "carrier",
    y = "mean delay (minutes)",
    title = "Mean departure delay by carrier",
    fill = "airport"
  ) +
  theme_bw() +
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9")) +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1)) +
  coord_flip()
```

# Which carrier has the most arrival delays?

```{r}
flights_dt %>% 
  filter(arr_delay > 0) %>% 
  group_by(carrier_name, origin) %>% 
  summarise(mean_delay = mean(arr_delay), .groups = "drop") %>% 
  ggplot() +
  aes(x = reorder(carrier_name, mean_delay), y = mean_delay, fill = origin) +
  geom_col(alpha = 0.8) +
  facet_wrap(~ origin) +
  labs(
    x = "carrier",
    y = "mean delay (minutes)",
    title = "Mean arrival delay by carrier",
    fill = "airport"
  ) +
  theme_bw() +
  scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9")) +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1)) +
  coord_flip()
```

# Which season has the highest number of departure delays?

```{r}
flights_dt %>% 
  filter(dep_delay > 0) %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") %>% 
  ggplot() +
  aes(x = day, y = mean_delay, colour = origin) +
  geom_point(alpha = 0.8) +
  labs(
    x = "date",
    y = "mean delay (minutes)",
    title = "Mean departure delay by month"
  ) +
  scale_colour_manual(values = c("#999999", "#E69F00", "#56B4E9")) +
  theme_bw()
```

# Over the course of a day when is the largest average delay?

```{r}
flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  select(dep_delay, hour) %>%
  group_by(hour) %>% 
  summarise(mean_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") %>% 
  ggplot() +
  aes(x = hour, y = mean_delay) +
  geom_point(alpha = 0.8, colour = "#56B4E9") +
  geom_smooth(se = FALSE, colour = "#E69F00") +
  labs(
    x = "hour",
    y = "mean delay (minutes)",
    title = "Mean departure delay by departure time"
  ) +
  theme_bw() +
  scale_x_continuous(limits = c(5, 23), breaks = seq(5, 23, by = 1))
```
# How does distance of an outbound flight affect departure delays?

```{r}
flights_dt %>% 
  filter(origin == "EWR") %>% 
  ggplot() +
  aes(x = distance) +
  geom_histogram(bins = 50) +
  theme_bw() +
  labs(title = "Distribution of flight distance")
```

```{r}
positions <- c("<500mi", "500-1000mi", "1000-1500mi", "1500-2000mi", "2000-2500mi", "2500-3000mi",
               ">3000mi")

flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  mutate(distance = case_when(
    distance <= 500 ~ "<500mi",
    distance > 500 & distance <= 1000 ~ "500-1000mi",
    distance > 1000 & distance <= 1500 ~ "1000-1500mi",
    distance > 1500 & distance <= 2000 ~ "1500-2000mi",
    distance > 2000 & distance <= 2500 ~ "2000-2500mi",
    distance > 2500 & distance <= 3000 ~ "2500-3000mi",
    distance > 3000 ~ ">3000mi"
  )) %>% 
  group_by(distance) %>% 
  summarise(mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = distance, y = mean_delay) +
  geom_col(fill = "#999999", alpha = 0.8) +
  labs(
    x = "distance (miles)",
    y = "mean departure delay (minutes)",
    title = "Departure delay by flight distance"
  ) +
  scale_x_discrete(limits = positions) +
  theme_bw()
```

# Which day is the best day to travel from Newark Int.?

```{r}
flights_dt %>%
  filter(dep_delay > 0 & origin == "EWR") %>% 
  mutate(weekday = wday(sched_dep_time, label = TRUE)) %>% 
  group_by(weekday) %>% 
  summarise(mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = weekday, y = mean_delay) +
  geom_col(fill = "#999999", alpha = 0.8) +
  labs(
    x = "weekday",
    y = "mean departure delay (minutes)",
    title = "Departure delay by weekday"
  ) +
  theme_bw()
```

# Which are the most popular destinations from Newark Int?

```{r}
flights_dt %>% 
  filter(origin == "EWR") %>% 
  group_by(dest, dest_name) %>% 
  summarise(count = n(), .groups = "drop") %>% 
  arrange(desc(count)) %>% 
  head(10) %>% 
  ggplot() +
  aes(x = reorder(dest_name, count), y = count) +
  geom_col(fill = "#999999", alpha = 0.8) +
  labs(
    x = "destination",
    y = "number of flights per year",
    title = "Destinations from Newark Int."
  ) +
  theme_bw() +
  coord_flip()
```

# Which destinations suffer from the longest delays?

```{r}
flights_dt %>%
  drop_na(dest_name) %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(dest, dest_name) %>% 
  summarise(mean_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") %>% 
  arrange(desc(mean_delay)) %>% 
  head(10) %>% 
  ggplot() +
  aes(x = reorder(dest_name, mean_delay), y = mean_delay) +
  geom_col(fill = "#999999", alpha = 0.8) +
  labs(
    x = "destination",
    y = "mean delay (minutes)",
    title = "Destinations with longest delays"
  ) +
  theme_bw() +
  coord_flip()
```

# Trend weather conditions against mean departure delays

```{r}
flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_visibility = mean(visib, na.rm = TRUE),
            mean_delay = mean(dep_delay), .groups = "drop") %>% 
  ggplot() +
  aes(x = mean_visibility, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean visibility (miles)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by visibility"
  ) +
  theme_bw()
```

```{r}
flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_wind_speed = mean(wind_speed, na.rm = TRUE),
            mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = mean_wind_speed, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean wind speed (mph)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by wind speed"
  ) +
  theme_bw()
```

```{r}
flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_wind_dir = mean(wind_dir, na.rm = TRUE),
            mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = mean_wind_dir, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean wind direction (degrees)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by wind direction"
  ) +
  theme_bw()
```

```{r}
flights_dt %>%
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_humidity = mean(humid, na.rm = TRUE),
            mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = mean_humidity, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean humidity (%)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by humidity"
  ) +
  theme_bw()
```

```{r}
flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_temp = mean(temp, na.rm = TRUE),
            mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = mean_temp, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean temperature (degf)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by temperature"
  ) +
  theme_bw()
```

```{r}
flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_dewpoint = mean(dewp, na.rm = TRUE),
            mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = mean_dewpoint, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean dewpoint (degF)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by dewpoint"
  ) +
  theme_bw()
```

```{r}
flights_dt %>% 
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_precip = mean(precip, na.rm = TRUE),
            mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = mean_precip, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean precipitation (inches)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by precipitation"
  ) +
  theme_bw()
```

```{r}
flights_dt %>%
  filter(dep_delay > 0 & origin == "EWR") %>% 
  group_by(day = floor_date(sched_dep_time, "day"), origin) %>% 
  summarise(mean_pressure = mean(pressure, na.rm = TRUE),
            mean_delay = mean(dep_delay)) %>% 
  ggplot() +
  aes(x = mean_pressure, y = mean_delay) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "#E69F00") +
  labs(
    x = "mean pressure (mbar)",
    y = "mean departure delay (minutes)",
    title = "Mean departure delay by pressure"
  ) +
  theme_bw()
```



