---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)

airlines <- read_csv("raw_data/airlines.csv")
airports <- read_csv("raw_data/airports.csv")
flights <- read_csv("raw_data/flights.csv")
places <- read_csv("raw_data/planes.csv")
weather <- read_csv("raw_data/weather.csv")
```

```{r}
summary(flights)

colnames(flights)
```

This crowdedness causes increasingly frequent delays in most major airports worldwide

The results show that adverse weather conditions, low ceilings, and low visibility conditions strongly influence flight delay

- Are longer flights more likely to be delayed or not?


- which airports are most susceptible to delays? 
- what is there geographical location?
- which is the busiest airport? time diff between landings
- A movement is a landing or takeoff of an aircraft.

- which routes are the most affected?
- which are the best routes?

- Only consider commercial airlines

- classify the weather dataset
- what conditions are worst for pilots?
- when is best to travel?


- which carriers are most susceptible to delays? 
- what does there fleet look like?
- domestic or international?
- which airplanes are most susceptible to delays?


- what weather leads to be most delays?
